Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models

### 2020-09-21 analysis/paper/paper.Rmd
made several edits based on Sarah's comments and response to reviewers.

```{r package-bibtex, eval = F}
knitr::write_bib(c("FSA", 
                   "Rcapture", 
                   "minerva", 
                   "quantregForest", 
                   "missForest",
                   "survey",
                   "sf", 
                   "knitr", 
                   "rmarkdown"),
                 file = 'packages.bib')
```

@@

The decline of anadromous Pacific salmonids (*Oncorhynchus spp.*) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend. These actions are often categorized into four H’s – harvest modification, hatchery practices, hydro-system operations, and habitat rehabilitation. Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality. The most recent “de-listing” criteria [@NOAA2016; @NOAA2016b] identified adult escapement targets at the population scale, providing a quantitative metric useful for evaluating the magnitude of survival improvements required. These abundance targets provide a benchmark against which habitat rehabilitation actions can be measured. Here we describe an approach for estimating life-stage specific habitat-based carrying capacity that can be used to quantitatively identify the magnitude of tributary habitat rehabilitation necessary to support de-listing. For perhaps the first time, the necessity of tributary habitat rehabilitation can be demonstrated and the magnitude of required change can be placed in context with the other “H’s.”

Pacific salmon species have experienced large declines in abundance throughout much of their range [@Good2005]. Declines can be partially attributed to lost or altered habitat, and thus, efforts to recover depleted salmon populations are replete with efforts to rehabilitate habitat used during the freshwater life-stages. Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*O. mykiss*) in the Pacific Northwest, USA [@NOAA2016; @NOAA2016b]. Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles. However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge. Reliable methods to better understand fish-habitat relationships and estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity to better direct tributary rehabilitation efforts.

@Fausch1988 conducted a thorough review of attempts to predict the abundance of fish from measurable habitat covariates from 1950 to 1985, and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal. Since that review, there has been progress in identifying some fish-habitat relationships for several salmonid species. @Nickelson1992 demonstrated that juvenile coho salmon (*O. kisutch*) were found in higher densities within pool habitat on the Oregon coast. Similarly, pool and pond densities were good predictors of coho smolt densities in western Washington [@Sharma2001]. @Bryant2009 found that juvenile coho abundance was positively related to large wood at the reach scale; however their results demonstrated a negative relationship between abundance and the number of pools. @Braun2011 similarly found positive associations between spawner densities of sockeye salmon (*O. nerka*) in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools. Densities of adult spawning coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed [@Pess2002]. @Mossop2006 examined juvenile Chinook salmon (*O. tshawytscha*) in the Yukon river and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient. These studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed. In addition, they all assumed some form of linear fish-habitat relationship which often results in weak predictive power.

@@

Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity. Further, many of these studies focus on only one or two measures of habitat. @Sweka2010 estimated carrying capacity of Atlantic salmon (*Salmo salar*) parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area. Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves; however, this requires a long time-series of data with variety in the number of spawners, data which are unavailable in most watersheds [@Cramer2009].

@@

Our study incorporates multiple measures of stream habitat to estimate fish-habitat relationships that encompass the collinear nature of most stream habitat metrics and can be used to predict carrying capacity. Our approach moves across several spatial scales, inferring fish-habitat relationships from detailed, localized habitat data and extrapolating capacity predictions across wide swaths of unsampled locations.

@@

After model fitting, the QRF model was then used to predict capacity at sites with measurements of the habitat covariates that were used to fit the model. In our case, this includes all sites within CHaMP basins in the interior Columbia River basin. For CHaMP sites that were sampled in multiple years, we first calculated the mean for each habitat metric among years to make predictions. In total, we generated `r filter(all_preds, is.na(chnk_per_m_se)) %>% summarise(n_sites = n_distinct(Site)) %>% pull(n_sites)` predictions of Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, South Fork Salmon, Tucannon, Wenatchee and Yankee Fork Salmon.

@@

A QRF model was fit using those metrics and the *quantregForest* package [@Meinshausen2006] in R [@Rsoftware2019] and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity. After model fit, we examined the relative importance of each habitat covariate included in the model (Figure \@ref(fig:rel-imp-figure)). Moreover, QRF models allow one to visually examine the marginal effect of each habitat covariate on the quantile of interest using partial dependence plots. These plots show the marginal effect of changing a single habitat covariate while maintaining all other covariates at their mean values (Figure \@ref(fig:pdp-figure)). However, given that many habitat metrics are somewhat correlated, these marginal effects are often not independent of one another and care should be taken when interpreting them. After model fitting, the QRF model was used to predict habitat capacity at all CHaMP sites within the interior Columbia River basin.

@@

## What Carrying Capacity Models Are Currently Available?

When it comes to estimating carrying capacity, spawner-recruit models are the gold standard [@Moussalli1986; @Myers1999]. However, such models require a long time-series of accurate estimates of adults and juveniles, with variation in the number of adults. Such datasets are not available everywhere, nor do they necessarily allow one to link capacity to habitat characteristics, except perhaps at the watershed scale. QRF models require a large number of paired fish-habitat sites, but if done across a wide variety of habitat types this data could theoretically be collected in a single (or very few) season(s). Or the dataset could be built over time, sampling different locations in different years. If the latter approach is taken, some thought must be given to possible year effects and how to best deal with them.

Historically, fish-habitat relationship models have focused on species distribution or average abundance/density rather than directly addressing carrying capacity. Bioenergics approaches, such as the net rate of energy intake (NREI) have been applied to salmonids to estimate capacity on the same reach scale that our QRF model operates on [@Wall2016]. However, there are some potential issues with how the food supply (i.e. invertebrate drift) is measured with these methods that could lead to biases in capacity estimates [@Dodrill2016] as well as difficulty in properly constraining drift depletion and inter-species competition, and computational and spatial limitation of this modelling approach [@Carmichael2020; @Wall2016]. The QRF approach described here allows us to analyze multiple (often correlated) habitat metrics and assess the potentially non-linear fish habitat relationships that define capacity as a function of habitat.

>>>>

```{r package-bibtex, eval = F}
knitr::write_bib(c("FSA", 
                   "Rcapture", 
                   "minerva", 
                   "quantregForest", 
                   "randomForest",
                   "missForest",
                   "survey",
                   "sf", 
                   "knitr", 
                   "rmarkdown"),
                 file = 'packages.bib')
```

@@

The decline of anadromous Pacific salmonids (*Oncorhynchus spp.*) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend. These actions are often categorized into four H’s – harvest modification, hatchery practices, hydro-system operations, and habitat rehabilitation. Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality. Freshwater habitat capacity deficits have recently been identified as a major factor directly impacting population abundance which has been largely overlooked in Columbia Basin salmonids [@Bond2018; @Hinrichsen2020; @NOAA2020]. Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*O. mykiss*) in the Pacific Northwest, USA [@NOAA2016; @NOAA2016b]. Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles. However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge [@Bond2018; @Hinrichsen2020; @NOAA2020]. Reliable methods to better understand fish-habitat relationships and estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity to better direct tributary rehabilitation efforts.

When it comes to estimating carrying capacity, spawner-recruit models are the gold standard [@Moussalli1986; @Myers1999]. However, such models require a long time-series of accurate estimates of adults and juveniles, with variation in the number of adults. Such data are unavailable in most watersheds [@Cramer2009], and they do not necessarily allow one to link capacity to habitat characteristics, except perhaps at the watershed scale. Bioenergetics approaches, such as the net rate of energy intake (NREI) have been applied to salmonids to estimate capacity on the 200 - 600 m reach scale [@Wall2016]. However, there are some potential issues with how the food supply (i.e. invertebrate drift) is measured with these methods that could lead to biases in capacity estimates [@Dodrill2016] as well as difficulty in properly constraining drift depletion and inter-species competition, and computational and spatial limitation of this modelling approach [@Wall2016; @Carmichael2020]. Taking a different approach, @Sweka2010 estimated carrying capacity of Atlantic salmon (*Salmo salar*) parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area.

Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity. Further, many of these studies focus on only one or two measures of habitat. @Fausch1988 conducted a thorough review of attempts to predict the abundance of fish from measurable habitat covariates from 1950 to 1985, and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal. Since that review, there has been progress in identifying some fish-habitat relationships for several salmonid species. @Nickelson1992 demonstrated that juvenile Coho salmon (*O. kisutch*) were found in higher densities within pool habitat on the Oregon coast. Similarly, pool and pond densities were good predictors of Coho smolt densities in western Washington [@Sharma2001]. @Bryant2009 found that juvenile Coho abundance was positively related to large wood at the reach scale; however their results demonstrated a negative relationship between abundance and the number of pools. @Braun2011 similarly found positive associations between spawner densities of sockeye salmon (*O. nerka*) in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools. Densities of adult spawning Coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed [@Pess2002]. @Mossop2006 examined juvenile Chinook salmon (*O. tshawytscha*) in the Yukon river and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient. These studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed. In addition, they all assumed some form of linear fish-habitat relationship which often results in weak predictive power.

@@

@@

Our study incorporates multiple measures of stream habitat to estimate fish-habitat relationships that encompass the collinear nature of most stream habitat metrics and can be used to predict carrying capacity. Our approach moves across several spatial scales, inferring fish-habitat relationships from detailed, localized habitat data and extrapolating capacity predictions across wide swaths of unsampled locations. Additionally, this approach for estimating life-stage specific habitat-based carrying capacity can be used to quantitatively identify the magnitude of tributary habitat rehabilitation necessary to support de-listing. Given the multitude of (often correlated) habitat metrics and the potentially non-linear fish habitat relationships that define capacity as a function of habitat, we explore the application of QRF modeling to habitat capacity estimation, validated using data from Columbia River Chinook salmon. For perhaps the first time, the necessity of tributary habitat rehabilitation can be demonstrated and the magnitude of required change can be placed in context with the other "H's."

@@

After model fitting, the QRF model was then used to predict capacity at sites with measurements of the habitat covariates that were used to fit the model. In our case, this includes all sites within CHaMP basins in the interior Columbia River basin. For CHaMP sites that were sampled in multiple years, we first calculated the mean for each habitat metric among years to make predictions. In total, we generated `r filter(all_preds, is.na(chnk_per_m_se)) %>% summarise(n_sites = n_distinct(Site)) %>% pull(n_sites)` predictions of Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, South Fork Salmon, Tucannon, Wenatchee and Yankee Fork Salmon. CHaMP sampled between 1 and 28% of the Chinook domain within these watersheds, with an average of 11%.

@@

A QRF model was fit using those metrics and the *quantregForest* package [@Meinshausen2006] in R [@Rsoftware2019] and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity. After model fit, we examined the relative importance of each habitat covariate included in the model (Figure \@ref(fig:rel-imp-figure)), quantified by the average decrease in residual sum of squares for splits on that variable amidst the trees in the random forest, implemented by the *importance* function from the *randomForest* package [@randomForest2002]. Moreover, QRF models allow one to visually examine the marginal effect of each habitat covariate on the quantile of interest using partial dependence plots. These plots show the marginal effect of changing a single habitat covariate while maintaining all other covariates at their mean values (Figure \@ref(fig:pdp-figure)). However, given that many habitat metrics are somewhat correlated, these marginal effects are often not independent of one another and care should be taken when interpreting them. After model fitting, the QRF model was used to predict habitat capacity at all CHaMP sites within the interior Columbia River basin.

@@

There are two aspects that make this approach "data hungry", meaning a large dataset is needed to fit a QRF model like this. First, random forest models generally require more data that parametric models, due to the lack of parametric distribution assumptions and the lack of an assumed form of the relationship between dependent and independent variables. Second, it takes larger data sets to accurately predict the lower and higher quantiles in a quantile regression framework. For example, if a data set consisted of thousands, rather than hundreds, of data points, a researcher might feel comfortable using the 95th or the 98th quantile as a proxy for capacity, rather than the 90th. Our data set consisted of `r n_distinct(qrf_data$Site)` sites, across a variety of habitats and years, providing contrast in all the habitat covariates and presumeably satisfying the data hungriness of a QRF model, based on our validation with spawner-recruit data.
